---
title: "Getting started with svyflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with svyflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

`svyflow` turns survey data into **design-correct descriptive statistics** —
proportions, means, totals, quantiles, min/max — with standard errors and
confidence intervals, driven either by an analysis plan (batch) or by
single-indicator wrappers (ad hoc). This vignette walks through a complete
workflow on a small synthetic household survey.

```{r}
library(svyflow)
```

## A synthetic dataset

We build a 500-household survey with a sampling weight, a stratification
variable (`province`), a couple of categorical questions, a multi-select
question stored as a delimited string, and two numeric variables.

```{r}
set.seed(123)
n <- 500

survey_df <- data.frame(
  province = sample(c("Kabul", "Balkh", "Herat"), n, replace = TRUE),
  gender   = sample(c("male", "female"), n, replace = TRUE, prob = c(0.49, 0.51)),
  edu_lvl  = sample(c("none", "primary", "secondary", "tertiary"), n, replace = TRUE),
  hh_size  = pmin(stats::rpois(n, lambda = 5) + 1L, 18L),
  income   = round(stats::rgamma(n, shape = 2, scale = 200)),
  weight   = stats::runif(n, 0.5, 2.0),
  stringsAsFactors = FALSE
)

# A multi-select "household needs" question, stored as "; "-delimited text.
needs <- c("cash", "food", "shelter", "health")
survey_df$hh_needs <- vapply(seq_len(n), function(i) {
  k <- sample(1:3, 1)
  paste(sort(sample(needs, k)), collapse = "; ")
}, character(1))

head(survey_df)
```

## Step 1 — handle the multi-select

Kobo / SurveyCTO usually export multi-select questions as one binary column
per option (`hh_needs/cash`, `hh_needs/food`, …). `svyflow` detects those
automatically. Here we only have the delimited string, so we expand it first:

```{r}
survey_df <- expand_multiselect(survey_df, vars = "hh_needs", sep = "; ")
attr(survey_df, "ms_options")$hh_needs
```

## Step 2 — build the survey design

`make_design()` wraps `survey::svydesign()` / `srvyr::as_survey()`. Pass the
column names as strings. `weights = NULL` gives simple random sampling, so the
same code path works for unweighted data.

```{r}
des <- make_design(survey_df, weights = "weight", strata = "province")
des
```

## Step 3 — batch analysis with a plan

An analysis plan is a data frame with one row per *(indicator x
disaggregation)*. The optional `variable_label` / `disaggregation_label`
columns produce friendly output names.

```{r}
plan <- tibble::tribble(
  ~variable,  ~kobo_type,        ~aggregation_method, ~disaggregation, ~variable_label,   ~disaggregation_label,
  "gender",   "select_one",      NA,                  "all",           "Gender",          NA,
  "edu_lvl",  "select_one",      NA,                  "gender",        "Education level", "Sex",
  "hh_needs", "select_multiple", NA,                  "all",           "Household needs", NA,
  "hh_size",  "integer",         "mean",              "gender",        "Household size",  "Sex",
  "income",   "integer",         "median",            "all",           "Monthly income",  NA
)

results <- analyze_survey(des, plan)
results
```

The result is a long tibble (class `svyflow_results`) with a stable schema:
`Disaggregation`, `Disaggregation_level`, `Question`, `Response`,
`Aggregation_method`, `Result`, `SE`, `CI_low`, `CI_high`, `Count`,
`Denominator`, `repeat_for`.

You can validate a plan on its own before running it:

```{r}
validate_plan(plan, survey_df)  # invisibly returns TRUE when valid
```

## Step 4 — choose how results are reported

Categorical estimates default to **proportions** (0–1). Switch the scale with
`result_format` and round with `digits`:

```{r}
res_pct <- analyze_survey(des, plan, result_format = "percent", digits = 1)
head(res_pct[res_pct$Aggregation_method == "perc", c("Question", "Response", "Result")])
```

Re-running the whole analysis just to change presentation is wasteful for big
plans. `format_results()` reformats an existing result cheaply, and is
aggregation-method aware (means/medians are never treated as percentages):

```{r}
format_results(results, to = "percent_fmt", digits = 1)
```

## Step 5 — single-indicator wrappers

For ad hoc work, the `summarize_*()` functions give a publication-ready table
for one indicator without writing a plan.

```{r}
summarize_select_one(des, "edu_lvl", variable_label = "Education level")
```

Disaggregated, pivoted to a **crosstab**:

```{r}
summarize_select_one(des, "edu_lvl",
                     disaggregation = "gender",
                     variable_label = "Education level",
                     disaggregation_label = "Sex",
                     crosstab = TRUE)
```

The same crosstab with confidence intervals inside each cell:

```{r}
summarize_select_one(des, "edu_lvl",
                     disaggregation = "gender",
                     variable_label = "Education level",
                     crosstab = TRUE, with_ci = TRUE, digits = 2)
```

Numeric statistics name the value column after the statistic:

```{r}
summarize_mean(des, "hh_size", variable_label = "Household size")
summarize_median(des, "income", variable_label = "Monthly income")
summarize_quantile(des, "income", q = 0.90, variable_label = "Monthly income")
```

Use `result_only = TRUE` for a compact one-line summary:

```{r}
summarize_mean(des, "hh_size", variable_label = "Household size",
               result_only = TRUE)
```

And multi-select:

```{r}
summarize_select_multiple(des, "hh_needs",
                          variable_label = "Household needs",
                          result_format = "percent", digits = 1)
```

## Step 6 — confidence-interval methods

By default `svyflow` reports a 95% t-interval on the design degrees of freedom
with plain Wald proportions. The `ci_opts()` bundle changes this and threads
through every function via the `ci` argument.

A 90% normal-approximation interval:

```{r}
summarize_mean(des, "hh_size", variable_label = "Household size",
               ci = ci_opts(ci_level = 0.90, df = Inf))
```

Logit intervals for proportions — preferable for **rare outcomes** near 0% or
100%, where the plain Wald interval can fall outside `[0, 1]`:

```{r}
summarize_select_one(des, "gender",
                     result_format = "percent",
                     ci = ci_opts(prop_method = "logit"))
```

In batch mode the method-specific knobs only affect the rows they apply to:
`prop_method` changes `select_one` / `select_multiple` rows, `interval_type` /
`qrule` change quantiles, and numeric rows are untouched.

## Step 7 — design effect and effective sample size

For methodology annexes, opt in to design-effect reporting with
`deff = TRUE`. The result gains a `DEFF` column (design effect) and an
`n_eff` column (effective sample size = `Denominator / DEFF`):

```{r}
res_deff <- analyze_survey(des, plan, deff = TRUE)
head(res_deff[res_deff$Aggregation_method %in% c("perc", "mean"),
              c("Question", "Response", "Result", "DEFF", "n_eff",
                "Denominator")])
```

DEFF compares the variance under your survey design to a simple-random
sample of the same size. Values around 1 indicate the design is as
efficient as SRS; values above 1 mean unequal weighting or intra-cluster
correlation is reducing precision. `n_eff` translates that back into an
SRS-equivalent sample size — useful for caveating reports built on
heavily clustered data.

DEFF is populated for `select_one`, `select_multiple`, `mean`, and `sum`
rows. Quantile, min and max rows carry `NA` because the underlying
`survey::svyquantile()` does not compute a design effect and extrema
bypass the survey design.

## Step 8 — export to Excel

`write_xlsx()` turns the result into a styled `.xlsx` workbook in one call.
The workbook layout is fixed: an `Overall` sheet first, one sheet per
disaggregation variable next, and a `Long` sheet at the end holding the
full long-form output as a flat table.

Add a `group` column to your plan to render section separators between
question blocks within each sheet:

```{r}
plan_grouped <- tibble::tribble(
  ~variable,  ~kobo_type,        ~aggregation_method, ~disaggregation, ~group,         ~variable_label,
  "gender",   "select_one",      NA,                  "all",           "Demographics", "Gender",
  "edu_lvl",  "select_one",      NA,                  "gender",        "Demographics", "Education level",
  "hh_size",  "integer",         "mean",              "gender",        "Household",    "Household size",
  "hh_needs", "select_multiple", NA,                  "gender",        "Needs",        "Household needs"
)
res_grouped <- analyze_survey(des, plan_grouped)

f <- tempfile(fileext = ".xlsx")
write_xlsx(res_grouped, f)
```

Display options on `write_xlsx()`:

```{r, eval = FALSE}
# Confidence intervals composed into each value cell.
write_xlsx(res_grouped, "with_ci.xlsx", with_ci = TRUE)

# Unweighted counts on the row labels (level totals).
write_xlsx(res_grouped, "with_counts.xlsx", with_counts = "row_label")

# Inline per-cell counts; combines with with_ci.
write_xlsx(res_grouped, "inline.xlsx",
           with_ci = TRUE, with_counts = "inline")

# A sibling "(n)" column after every value column.
write_xlsx(res_grouped, "parallel.xlsx", with_counts = "parallel")
```

Because the exporter writes values verbatim, control the display
format upstream with `format_results()`:

```{r, eval = FALSE}
res_pct <- format_results(res_grouped, to = "percent_fmt", digits = 1)
write_xlsx(res_pct, "report.xlsx", with_ci = TRUE)
```

Restyle with `xlsx_theme()` — override any field (font, header fill,
borders, section styling):

```{r, eval = FALSE}
write_xlsx(res_grouped, "navy.xlsx",
           theme = xlsx_theme(font_name = "Arial",
                              header_fill = "#1E3A5F"))
```

Value-column width is fixed (`col_width = 21` by default); long headers
wrap within that width. The row-label column sizes itself to content.

## Step 9 — methods / reproducibility sheet

`write_methods_xlsx()` is a **standalone companion** to `write_xlsx()`:
it writes a single-sheet workbook documenting how the analysis was
produced — session info, the design, sample sizes, weights, CI options,
and (when supplied) a plan summary and a DEFF roll-up. Ship it alongside
the results workbook so a recipient can interpret and reproduce them.

```{r, eval = FALSE}
write_methods_xlsx(
  "methods.xlsx",
  design  = des,
  plan    = plan,
  results = res_grouped,
  ci      = ci_opts(),
  cover_notes = c(Project = "2026 baseline",
                  Funder  = "Donor agency",
                  "Estimates use design-correct standard errors.")
)
```

The function never touches `analyze_survey()` or `write_xlsx()`. The
`plan`, `results` and `cover_notes` arguments are all optional — supply
only what you have. When `results` carries a `DEFF` column, a roll-up
section (mean / median / max DEFF, mean `n_eff`, top 5 highest DEFFs) is
added automatically.

## Step 10 — group comparison

Descriptives answer "what is the value?"; `compare_groups()` answers "is the
difference real?". It selects and runs the right survey-design-aware test for
each indicator across the levels of a grouping variable and returns one tidy
row per indicator, with the test statistic, degrees of freedom, an effect
size, the p-value and significance stars.

```{r}
# Numeric indicators x two groups -> independent t-tests, auto-selected.
compare_groups(des, c("hh_size", "income"), "gender")

# Categorical indicator -> Rao-Scott chi-square (Fisher's exact on sparse
# tables), with Cramer's V as the effect size.
compare_groups(des, "edu_lvl", "province")

# Rank-based variants when normality is doubtful.
compare_groups(des, "hh_size", "province", parametric = FALSE)  # Kruskal-Wallis
```

For a **paired** design — the same units measured twice (baseline vs endline)
— supply `paired = TRUE` and the `pair_by` identifier so the test is run on
the per-unit differences:

```{r}
# A small synthetic two-wave panel.
set.seed(11)
n_units <- 150
base_v  <- stats::rpois(n_units, 5) + 1
panel <- data.frame(
  unit_id = rep(seq_len(n_units), each = 2),
  wave    = rep(c("baseline", "endline"), times = n_units),
  score   = c(rbind(base_v, base_v + stats::rnorm(n_units, 0.7, 1.5))),
  weight  = rep(stats::runif(n_units, 0.5, 2), each = 2)
)
pdes <- make_design(panel, weights = "weight")

compare_groups(pdes, "score", "wave", paired = TRUE, pair_by = "unit_id")
```

Any branch can be forced with `test =` (`"ttest"`, `"wilcoxon"`, `"anova"`,
`"kruskal"`, `"chisq"`, `"fisher"`, `"prop_z"`, plus the paired variants), and
`ci_opts()` threads the confidence level / degrees of freedom into the t-test
intervals. All tests except Fisher's exact respect the survey design; Fisher
has no survey-aware form, so it runs on the unweighted table and emits a note.

## Where to go next

- `?analyze_survey` — the batch entry point and its full plan schema.
- `?summarize_select_one`, `?summarize_mean`, … — the single-indicator family.
- `?compare_groups` — group comparison / significance testing.
- `?ci_opts` — confidence-interval options.
- `?format_results` — post-run reformatting.
- `?write_xlsx`, `?xlsx_theme` — Excel export and theming.
- `?make_design`, `?expand_multiselect` — design construction and multi-select.
