---
title: "Getting started with svyflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with svyflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

`svyflow` turns survey data into **design-correct descriptive statistics** —
proportions, means, totals, quantiles, min/max — with standard errors and
confidence intervals, driven either by an analysis plan (batch) or by
single-indicator wrappers (ad hoc). This vignette walks through a complete
workflow on a small synthetic household survey.

```{r}
library(svyflow)
```

## A synthetic dataset

We build a 500-household survey with a sampling weight, a stratification
variable (`province`), a couple of categorical questions, a multi-select
question stored as a delimited string, and two numeric variables.

```{r}
set.seed(123)
n <- 500

survey_df <- data.frame(
  province = sample(c("Kabul", "Balkh", "Herat"), n, replace = TRUE),
  gender   = sample(c("male", "female"), n, replace = TRUE, prob = c(0.49, 0.51)),
  edu_lvl  = sample(c("none", "primary", "secondary", "tertiary"), n, replace = TRUE),
  hh_size  = pmin(stats::rpois(n, lambda = 5) + 1L, 18L),
  income   = round(stats::rgamma(n, shape = 2, scale = 200)),
  weight   = stats::runif(n, 0.5, 2.0),
  stringsAsFactors = FALSE
)

# A multi-select "household needs" question, stored as "; "-delimited text.
needs <- c("cash", "food", "shelter", "health")
survey_df$hh_needs <- vapply(seq_len(n), function(i) {
  k <- sample(1:3, 1)
  paste(sort(sample(needs, k)), collapse = "; ")
}, character(1))

head(survey_df)
```

## Step 1 — handle the multi-select

Kobo / SurveyCTO usually export multi-select questions as one binary column
per option (`hh_needs/cash`, `hh_needs/food`, …). `svyflow` detects those
automatically. Here we only have the delimited string, so we expand it first:

```{r}
survey_df <- expand_multiselect(survey_df, vars = "hh_needs", sep = "; ")
attr(survey_df, "ms_options")$hh_needs
```

## Step 2 — build the survey design

`make_design()` wraps `survey::svydesign()` / `srvyr::as_survey()`. Pass the
column names as strings. `weights = NULL` gives simple random sampling, so the
same code path works for unweighted data.

```{r}
des <- make_design(survey_df, weights = "weight", strata = "province")
des
```

## Step 3 — batch analysis with a plan

An analysis plan is a data frame with one row per *(indicator x
disaggregation)*. The optional `variable_label` / `disaggregation_label`
columns produce friendly output names.

```{r}
plan <- tibble::tribble(
  ~variable,  ~kobo_type,        ~aggregation_method, ~disaggregation, ~variable_label,   ~disaggregation_label,
  "gender",   "select_one",      NA,                  "all",           "Gender",          NA,
  "edu_lvl",  "select_one",      NA,                  "gender",        "Education level", "Sex",
  "hh_needs", "select_multiple", NA,                  "all",           "Household needs", NA,
  "hh_size",  "integer",         "mean",              "gender",        "Household size",  "Sex",
  "income",   "integer",         "median",            "all",           "Monthly income",  NA
)

results <- analyze_survey(des, plan)
results
```

The result is a long tibble (class `svyflow_results`) with a stable schema:
`Disaggregation`, `Disaggregation_level`, `Question`, `Response`,
`Aggregation_method`, `Result`, `SE`, `CI_low`, `CI_high`, `Count`,
`Denominator`, `repeat_for`.

You can validate a plan on its own before running it:

```{r}
validate_plan(plan, survey_df)  # invisibly returns TRUE when valid
```

## Step 4 — choose how results are reported

Categorical estimates default to **proportions** (0–1). Switch the scale with
`result_format` and round with `digits`:

```{r}
res_pct <- analyze_survey(des, plan, result_format = "percent", digits = 1)
head(res_pct[res_pct$Aggregation_method == "perc", c("Question", "Response", "Result")])
```

Re-running the whole analysis just to change presentation is wasteful for big
plans. `format_results()` reformats an existing result cheaply, and is
aggregation-method aware (means/medians are never treated as percentages):

```{r}
format_results(results, to = "percent_fmt", digits = 1)
```

## Step 5 — single-indicator wrappers

For ad hoc work, the `summarize_*()` functions give a publication-ready table
for one indicator without writing a plan.

```{r}
summarize_select_one(des, "edu_lvl", variable_label = "Education level")
```

Disaggregated, pivoted to a **crosstab**:

```{r}
summarize_select_one(des, "edu_lvl",
                     disaggregation = "gender",
                     variable_label = "Education level",
                     disaggregation_label = "Sex",
                     crosstab = TRUE)
```

The same crosstab with confidence intervals inside each cell:

```{r}
summarize_select_one(des, "edu_lvl",
                     disaggregation = "gender",
                     variable_label = "Education level",
                     crosstab = TRUE, with_ci = TRUE, digits = 2)
```

Numeric statistics name the value column after the statistic:

```{r}
summarize_mean(des, "hh_size", variable_label = "Household size")
summarize_median(des, "income", variable_label = "Monthly income")
summarize_quantile(des, "income", q = 0.90, variable_label = "Monthly income")
```

Use `result_only = TRUE` for a compact one-line summary:

```{r}
summarize_mean(des, "hh_size", variable_label = "Household size",
               result_only = TRUE)
```

And multi-select:

```{r}
summarize_select_multiple(des, "hh_needs",
                          variable_label = "Household needs",
                          result_format = "percent", digits = 1)
```

## Step 6 — confidence-interval methods

By default `svyflow` reports a 95% t-interval on the design degrees of freedom
with plain Wald proportions. The `ci_opts()` bundle changes this and threads
through every function via the `ci` argument.

A 90% normal-approximation interval:

```{r}
summarize_mean(des, "hh_size", variable_label = "Household size",
               ci = ci_opts(ci_level = 0.90, df = Inf))
```

Logit intervals for proportions — preferable for **rare outcomes** near 0% or
100%, where the plain Wald interval can fall outside `[0, 1]`:

```{r}
summarize_select_one(des, "gender",
                     result_format = "percent",
                     ci = ci_opts(prop_method = "logit"))
```

In batch mode the method-specific knobs only affect the rows they apply to:
`prop_method` changes `select_one` / `select_multiple` rows, `interval_type` /
`qrule` change quantiles, and numeric rows are untouched.

## Step 7 — export to Excel

`write_xlsx()` turns the result into a styled `.xlsx` workbook in one call.
The workbook layout is fixed: an `Overall` sheet first, one sheet per
disaggregation variable next, and a `Long` sheet at the end holding the
full long-form output as a flat table.

Add a `group` column to your plan to render section separators between
question blocks within each sheet:

```{r}
plan_grouped <- tibble::tribble(
  ~variable,  ~kobo_type,        ~aggregation_method, ~disaggregation, ~group,         ~variable_label,
  "gender",   "select_one",      NA,                  "all",           "Demographics", "Gender",
  "edu_lvl",  "select_one",      NA,                  "gender",        "Demographics", "Education level",
  "hh_size",  "integer",         "mean",              "gender",        "Household",    "Household size",
  "hh_needs", "select_multiple", NA,                  "gender",        "Needs",        "Household needs"
)
res_grouped <- analyze_survey(des, plan_grouped)

f <- tempfile(fileext = ".xlsx")
write_xlsx(res_grouped, f)
```

Display options on `write_xlsx()`:

```{r, eval = FALSE}
# Confidence intervals composed into each value cell.
write_xlsx(res_grouped, "with_ci.xlsx", with_ci = TRUE)

# Unweighted counts on the row labels (level totals).
write_xlsx(res_grouped, "with_counts.xlsx", with_counts = "row_label")

# Inline per-cell counts; combines with with_ci.
write_xlsx(res_grouped, "inline.xlsx",
           with_ci = TRUE, with_counts = "inline")

# A sibling "(n)" column after every value column.
write_xlsx(res_grouped, "parallel.xlsx", with_counts = "parallel")
```

Because the exporter writes values verbatim, control the display
format upstream with `format_results()`:

```{r, eval = FALSE}
res_pct <- format_results(res_grouped, to = "percent_fmt", digits = 1)
write_xlsx(res_pct, "report.xlsx", with_ci = TRUE)
```

Restyle with `xlsx_theme()` — override any field (font, header fill,
borders, section styling):

```{r, eval = FALSE}
write_xlsx(res_grouped, "navy.xlsx",
           theme = xlsx_theme(font_name = "Arial",
                              header_fill = "#1E3A5F"))
```

Value-column width is fixed (`col_width = 21` by default); long headers
wrap within that width. The row-label column sizes itself to content.

## Where to go next

- `?analyze_survey` — the batch entry point and its full plan schema.
- `?summarize_select_one`, `?summarize_mean`, … — the single-indicator family.
- `?ci_opts` — confidence-interval options.
- `?format_results` — post-run reformatting.
- `?write_xlsx`, `?xlsx_theme` — Excel export and theming.
- `?make_design`, `?expand_multiselect` — design construction and multi-select.
